The Impact of Anthropogenic Forcing on ENSO Amplitude

### 2021-10-31 paperwork/abstract/abstract.tex
more stuff woohoo

% Created 2021-05-10 Mon 19:08
% Intended LaTeX compiler: pdflatex
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\date{\today}
\title{Abstract}
\hypersetup{
 pdfauthor={John Doe},
 pdftitle={Abstract},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 27.2 (Org mode 9.5)}, 
 pdflang={English}}
\begin{document}

\begin{center}

{\bf The Impact of Industrial Emissions on ENSO Amplitude}
{\bf Benjamin Goldman}

\end{center}

The El Niño/Southern Oscillation (ENSO) is the dominant mode of interannual climate variability, with substantial associated global socio-economic impacts. Due to their significance, shifts in ENSO under climate change also have the potential to substantial impact human society and natural ecosystems. However, it is currently unclear what effect greenhouse gas (GHG) and industrial aerosol (AER) emissions will have on ENSO and even less clear what effect a combination of these factors might have when changing in tandem. This study examines transient changes to ENSO variance under a variety of forcing scenarios using the CESM1 Large and Single-Forcing Ensembles. These multi-member ensembles span the historical record (1920-2005) and much of the 21st C (2006-2080 for GHG/AER). A 2000-year pre-industrial (PI) control simulation is used to account for model drift and 20-year running variance of the Niño 3.4 SST index is used as a proxy for ENSO variance. The ensemble mean and standard error of each ensemble is calculated, to estimate the statistical significance of simulated changes. We identify significant increases in variance under full-forcing conditions during the historical record and attribute these mainly to changes in GHG, with the potential emergence of AER-driven increases in the decades to come. Computing detrended and smoothed correlation between 20-year variance of the Niño 3.4 index and ocean temperature in the tropical Pacific Ocean reveals that stratification caused by global warming is connected with increased ENSO amplitude.
\end{document}

>>>>

% Created 2021-10-28 Thu 18:11
% Intended LaTeX compiler: pdflatex
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\date{\today}
\title{Abstract}
\hypersetup{
 pdfauthor={Benjamin Goldman},
 pdftitle={Abstract},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 27.2 (Org mode 9.5)}, 
 pdflang={English}}
\begin{document}

\begin{center}
\textbf{The Impact of Industrial Emissions on ENSO Amplitude}

\textbf{Benjamin Goldman}
\end{center}

The El Niño/Southern Oscillation (ENSO) is the dominant mode of interannual climate variability, with substantial associated global socio-economic impacts. Due to their significance, shifts in ENSO under climate change also have the potential to substantially impact human society and natural ecosystems. However, it is currently unclear what effect greenhouse gas and industrial aerosol emissions have on ENSO, as well as what effect these factors have when combined. This study examined transient changes to ENSO variance under a variety of forcing scenarios using the CESM1 and CESM2 Large and Single-Forcing Ensembles. These multi-member ensembles span the historical record (1920-2005) and much of the 21st century (2006-2080 for GHG/AER). A 2000-year pre-industrial (PI) control simulation is used to account for model drift and 20-year running variance of the Niño 3.4 SST index is used as a proxy for ENSO variance. The ensemble mean and standard error of each ensemble was calculated, while the Probability Density Function (PDF) is computed for the PI control simulation to estimate the statistical significance of simulated changes. We calculated the correlation coefficient between ocean temperature in the equatorial Pacific and Niño 3.4 under various forcing conditions, concluding that Pacific stratification likely is tied to changes to ENSO amplitude. Finally, we examined the wavelet power spectrum of the historical and predicted Niño 3.4 index to support the data given from the 20-year variance calculation. We identified significant increases in variance of the Niño 3.4 index under full-forcing conditions during the historical record and attribute these mainly to changes in GHG, with the potential emergence of AER-driven increases in the decades to come.

\begin{center}
\textbf{Presentations}
\end{center}

\textbf{2020:} \emph{Virtual Poster Board presenter at the White Plains High School Science Research Symposium (June)}

\textbf{2021:} \emph{Virtual Powerpoint presenter at the Westchester Science and Engineering Fair (March)}

\begin{center}
\textbf{Awards}
\end{center}

\textbf{2021:} \emph{American Meteorological Society Award at the Westchester Science and Engineering Fair}

\textbf{2021:} \emph{4\textsuperscript{th} Place in Earth Science at the Westchester Science and Engineering Fair}

\begin{center}
\textbf{Mentor Information}

Dr. John Fasullo

National Center for Atmospheric Research

Boulder, Colorado
\end{center}
\end{document}

### 2021-10-31 paperwork/paper/paper.tex
more stuff woohoo

% Created 2021-10-14 Thu 09:16
% Intended LaTeX compiler: pdflatex
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{natbib}
\author{Benjamin Goldman}
\date{\today}
\title{The Impact of Anthropogenic Forcing on ENSO Amplitude}
\hypersetup{
 pdfauthor={Benjamin Goldman},
 pdftitle={The Impact of Anthropogenic Forcing on ENSO Amplitude},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 27.2 (Org mode 9.5)}, 
 pdflang={English}}
\begin{document}

\maketitle
\tableofcontents

\begin{abstract}
The El Niño/Southern Oscillation (ENSO) is the dominant mode of interannual climate variability, with substantial associated global socio-economic impacts. Due to their significance, shifts in ENSO under climate change also have the potential to substantially impact human society and natural ecosystems. However, it is currently unclear what effect greenhouse gas (GHG) and industrial aerosol (AER) emissions have on ENSO, as well as what effect these factors have when combined. This study examined transient changes to ENSO variance under a variety of forcing scenarios using the CESM1 and CESM2 Large and Single-Forcing Ensembles. These multi-member ensembles span the historical record (1920-2005) and much of the 21st C (2006-2080 for GHG/AER). A 2000-year pre-industrial (PI) control simulation is used to account for model drift and 20-year running variance of the Niño 3.4 SST index is used as a proxy for ENSO variance. The ensemble mean and standard error of each ensemble was calculated, while the Probability Density Function (PDF) is computed for the PI control simulation to estimate the statistical significance of simulated changes. We calculated the correlation coefficient between ocean temperature in the equatorial Pacific and Niño 3.4 under various forcing conditions, concluding that Pacific stratification likely is tied to changes to ENSO amplitude. We identifed significant increases in variance of the Niño 3.4 index under full-forcing conditions during the historical record and attribute these mainly to changes in GHG, with the potential emergence of AER-driven increases in the decades to come.
\end{abstract}

\section{Introduction}
\label{sec:org2a498ff}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{../../data/figures/ff_compare.pdf}
\caption{\label{fig:variance_2}20-year windowed variance of the Niño 3.4 index for the CESM1 and CESM2 fully-forced Large Ensembles. Colored bars represent ensemble standard error.}
\end{figure}

The fully-forced ensemble exhibits reduced variance in the mid-late 20\textsuperscript{th} century, below that of the PI control. The most likely explanation for this phenomenon is internal variability. To test the likelihood of this explanation, we test the influence of the Atlantic Multidecadal Oscillation (AMO) and Atlantic Meridional Overturning Current (AMOC) on ENSO in the control simulation. The AMO has been shown to have an influence on ENSO strength and seasonal growth rate \citep{levine2017impact}. We filter the control Niño 3.4 variance data according to the strength of the AMO and AMOC, using records of AMO/C strength in the Climate Variability Diognostics Package (CVDP) \citep{phillips2014evaluating}. The Probability Distribution Function of Niño 3.4 variance is estimated for AMO/C > 1/2\(\sigma\) and < -1/-2\(\sigma\) using a Kernel Density Estimation. No meaningful differences were found in the distribution of ENSO intensity under any of these conditions (See Suplementary information figure 1).

\subsection{Single Forcing Scenarios and the Bootstrap Process}
\label{sec:org75cad93}
We analyze the role of individual factors using the CESM Single Forcing Ensembles. To separate the influence of a single factor from the fully-forced ensemble, we employ a bootstrap test. For each single-forcing ensemble, a single simulation is randomly selected, and the Niño 3.4 20-year variance record is subtracted from that from a randomly selected fully-forced simulation. We repeat this process 1000 times for each ensemble, and then calculate the mean and standard error for each ensemble. These results are shown in figure \ref{fig:bootstrap_1}. The greenhouse-only ensemble as well as the aerosol-only ensemble exhibit increased variance, signaling that greenhouse and aerosol emissions likely both play a significant role in ENSO's forced response in the full-forcing ensemble. Interestingly, the influence of greenhouse gasses and aerosols are non-conflicting, in contrast with previous studies that show opposite effects of greenhouse gas and aerosol forcing on ENSO \citep{stevenson2017forced}. All the other single forcing ensembles exhibit insignificant differences from the fully-forced ensemble. The biomass burning case shows very small deviations from the fully-forced case, while the ozone ensemble's period of recording is too small to draw meaningful conclusions. However, \citet{nowack2017role} showed that ozone forcing may dampen the effect of greenhouse-forced increases to ENSO amplitude by reducing changes to Pacific sea temperature and the Walker Circulation. The land use/cover case, while it does show large deviations from the fully-forced case, has an ensemble size (5 members) too small to lend any credibility to these changes.

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{../../data/figures/cesm1_sf.pdf}
\caption{\label{fig:bootstrap_1}Difference between fully forced and single forced ensembles derived from the bootstrap process for a) aerosol emissions, b) biomass burning c) greenhouse gasses, d) land use/cover}
\end{figure}

@@

\subsection{Correlation With Ocean Temperature}
\label{sec:org5c7fa6f}
Next, we analyzeed the correlation between ENSO amplitude and changes to ocean structure in the CESM1 ensembles. To do this we used 4 slices of ocean temperature from the fully-forced, xghg, and xaer ensembles, including a slice averaged along the equator and slices through the western, central, and eastern Pacific basins. We linearly detrend and smooth with a 30-year windowed mean the timeseries at each gridpoint and the Niño 3.4 variance. Next, we calculated the Pearson's correlation coefficient between each gridpoint and the Niño 3.4 variance timeseries. The correlation coefficients for the equatorial slice are shown in figure \ref{fig:tempdt}.

>>>>

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{natbib}
\author{Benjamin Goldman}
\date{\today}
\title{The Impact of Anthropogenic Forcing on ENSO Amplitude}

\begin{document}

\maketitle

\begin{abstract}
The El Niño/Southern Oscillation (ENSO) is the dominant mode of interannual climate variability, with substantial associated global socio-economic impacts. Due to their significance, shifts in ENSO under climate change also have the potential to substantially impact human society and natural ecosystems. However, it is currently unclear what effect greenhouse gas and industrial aerosol emissions have on ENSO, as well as what effect these factors have when combined. This study examined transient changes to ENSO variance under a variety of forcing scenarios using the CESM1 and CESM2 Large and Single-Forcing Ensembles. These multi-member ensembles span the historical record (1920-2005) and much of the 21st century (2006-2080 for GHG/AER). A 2000-year pre-industrial (PI) control simulation is used to account for model drift and 20-year running variance of the Niño 3.4 SST index is used as a proxy for ENSO variance. The ensemble mean and standard error of each ensemble was calculated, while the Probability Density Function (PDF) is computed for the PI control simulation to estimate the statistical significance of simulated changes. We calculated the correlation coefficient between ocean temperature in the equatorial Pacific and Niño 3.4 under various forcing conditions, concluding that Pacific stratification likely is tied to changes to ENSO amplitude. Finally, we examined the wavelet power spectrum of the historical and predicted Niño 3.4 index to support the data given from the 20-year variance calculation. We identified significant increases in variance of the Niño 3.4 index under full-forcing conditions during the historical record and attribute these mainly to changes in GHG, with the potential emergence of AER-driven increases in the decades to come.
\end{abstract}

\section{Introduction}

@@

The fully-forced ensemble exhibits reduced variance in the mid-late 20\textsuperscript{th} century, below that of the PI control. The most likely explanation for this phenomenon is internal variability. To test the likelihood of this explanation, we test the influence of the Atlantic Multidecadal Oscillation (AMO) and Atlantic Meridional Overturning Current (AMOC) on ENSO in the control simulation. The AMO has been shown to have an influence on ENSO strength and seasonal growth rate \citep{levine2017impact}. We filter the control Niño 3.4 variance data according to the strength of the AMO and AMOC, using records of AMO/C strength in the Climate Variability Diognostics Package (CVDP) \citep{phillips2014evaluating}. The Probability Distribution Function of Niño 3.4 variance is estimated for AMO/C > 1/2\(\sigma\) and < -1/-2\(\sigma\) using a Kernel Density Estimation. No meaningful differences were found in the distribution of ENSO intensity under any of these conditions (See Supplementary information figure 1).

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{../../data/figures/cesm1.pdf}
\caption{\label{fig:variance_1}20-year windowed variance of the Niño 3.4 index, for ensembles a) full-forcing, b) xghg, c) xaer, d) xbmb, e) xlulc. Colored bars represent ensemble standard error.}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{../../data/figures/ff_compare.pdf}
\caption{\label{fig:variance_2}20-year windowed variance of the Niño 3.4 index for the CESM1 and CESM2 fully-forced Large Ensembles. Colored bars represent ensemble standard error.}
\end{figure}

\subsection{Single Forcing Scenarios and the Bootstrap Process}

We analyze the role of individual factors using the CESM Single Forcing Ensembles. To separate the influence of a single factor from the fully-forced ensemble, we employ a bootstrap test. For each single-forcing ensemble, a single simulation is randomly selected, and the Niño 3.4 20-year variance record is subtracted from that from a randomly selected fully-forced simulation. We repeat this process 1000 times for each ensemble, and then calculate the mean and standard error for each ensemble. These results are shown in figure \ref{fig:bootstrap_1}. The greenhouse-only ensemble as well as the aerosol-only ensemble exhibit increased variance, signaling that greenhouse and aerosol emissions likely both play a significant role in ENSO's forced response in the full-forcing ensemble. Interestingly, the influence of greenhouse gasses and aerosols are non-conflicting, in contrast with previous studies that show opposite effects of greenhouse gas and aerosol forcing on ENSO \citep{stevenson2017forced}. All the other single forcing ensembles exhibit insignificant differences from the fully-forced ensemble. The biomass burning case shows very small deviations from the fully-forced case, while the ozone ensemble's period of recording is too small to draw meaningful conclusions. However, \citet{nowack2017role} showed that ozone forcing may dampen the effect of greenhouse-forced increases to ENSO amplitude by reducing changes to Pacific sea temperature and the Walker Circulation. The land use/cover case, while it does show large deviations from the fully-forced case, has an ensemble size (5 members) too small to lend any credibility to these changes.

@@

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{../../data/figures/cesm1_sf.pdf}
\caption{\label{fig:bootstrap_1}Difference between fully forced and single forced ensembles derived from the bootstrap process for a) aerosol emissions, b) biomass burning c) greenhouse gasses, d) land use/cover}
\end{figure}

\subsection{Correlation With Ocean Temperature}

Next, we analyzed the correlation between ENSO amplitude and changes to ocean structure in the CESM1 ensembles. To do this we used 4 slices of ocean temperature from the fully-forced, xghg, and xaer ensembles, including a slice averaged along the equator and slices through the western, central, and eastern Pacific basins. We linearly detrend and smooth with a 30-year windowed mean the timeseries at each grid-point and the Niño 3.4 variance. Next, we calculated the Pearson's correlation coefficient between each grid-point and the Niño 3.4 variance timeseries. The correlation coefficients for the equatorial slice are shown in figure \ref{fig:tempdt}.

### 2021-10-31 paperwork/paper_old/paper.tex
more stuff woohoo

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{natbib}
\author{Benjamin Goldman}
\date{\today}
\title{The Impact of Anthropogenic Forcing on ENSO Amplitude}

\begin{document}

\maketitle

\begin{abstract}
The El Niño/Southern Oscillation (ENSO) is the dominant mode of interannual climate variability, with substantial associated global socio-economic impacts. Due to their significance, shifts in ENSO under climate change also have the potential to substantially impact human society and natural ecosystems. However, it is currently unclear what effect greenhouse gas and industrial aerosol emissions have on ENSO, as well as what effect these factors have when combined. This study examined transient changes to ENSO variance under a variety of forcing scenarios using the CESM1 and CESM2 Large and Single-Forcing Ensembles. These multi-member ensembles span the historical record (1920-2005) and much of the 21st century (2006-2080 for GHG/AER). A 2000-year pre-industrial (PI) control simulation is used to account for model drift and 20-year running variance of the Niño 3.4 SST index is used as a proxy for ENSO variance. The ensemble mean and standard error of each ensemble was calculated, while the Probability Density Function (PDF) is computed for the PI control simulation to estimate the statistical significance of simulated changes. We calculated the correlation coefficient between ocean temperature in the equatorial Pacific and Niño 3.4 under various forcing conditions, concluding that Pacific stratification likely is tied to changes to ENSO amplitude. Finally, we examined the wavelet power spectrum of the historical and predicted Niño 3.4 index to support the data given from the 20-year variance calculation. We identified significant increases in variance of the Niño 3.4 index under full-forcing conditions during the historical record and attribute these mainly to changes in GHG, with the potential emergence of AER-driven increases in the decades to come.
\end{abstract}

\section{Introduction}

@@

\section{Data}

@@

\section{Methods and Results}

\subsection{Niño 3.4 Variance}

We estimate 20\textsuperscript{th} and 21\textsuperscript{st} century ENSO amplitude for each model in the fully-forced, single-forced, and PI control CESM1 and CESM2 ensembles using the variance of the Niño 3.4 region of the Pacific ocean (5N-5S, 170W-120W). First, we subtract the ensemble mean from each member to isolate the internal variability for each member. We measure variance on 20-year centered sliding windows. We calculate the multi-model mean for each ensemble, as well as the ensemble standard error. The results of this calculation  for the CESM1 Large and Single Forcing Ensembles are shown in figure \ref{fig:variance_1}.

@@

@@

The fully-forced ensemble exhibits reduced variance in the mid-late 20\textsuperscript{th} century, below that of the PI control. The most likely explanation for this phenomenon is internal variability. To test the likelihood of this explanation, we test the influence of the Atlantic Multidecadal Oscillation (AMO) and Atlantic Meridional Overturning Current (AMOC) on ENSO in the control simulation. The AMO has been shown to have an influence on ENSO strength and seasonal growth rate \citep{levine2017impact}. We filter the control Niño 3.4 variance data according to the strength of the AMO and AMOC, using records of AMO/C strength in the Climate Variability Diognostics Package (CVDP) \citep{phillips2014evaluating}. The Probability Distribution Function of Niño 3.4 variance is estimated for AMO/C > 1/2\(\sigma\) and < -1/-2\(\sigma\) using a Kernel Density Estimation. No meaningful differences were found in the distribution of ENSO intensity under any of these conditions (See Supplementary information figure 1).

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{../../data/figures/cesm1.pdf}
\caption{\label{fig:variance_1}20-year windowed variance of the Niño 3.4 index, for ensembles a) full-forcing, b) xghg, c) xaer, d) xbmb, e) xlulc. Colored bars represent ensemble standard error.}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{../../data/figures/ff_compare.pdf}
\caption{\label{fig:variance_2}20-year windowed variance of the Niño 3.4 index for the CESM1 and CESM2 fully-forced Large Ensembles. Colored bars represent ensemble standard error.}
\end{figure}

\subsection{Single Forcing Scenarios and the Bootstrap Process}

We analyze the role of individual factors using the CESM Single Forcing Ensembles. To separate the influence of a single factor from the fully-forced ensemble, we employ a bootstrap test. For each single-forcing ensemble, a single simulation is randomly selected, and the Niño 3.4 20-year variance record is subtracted from that from a randomly selected fully-forced simulation. We repeat this process 1000 times for each ensemble, and then calculate the mean and standard error for each ensemble. These results are shown in figure \ref{fig:bootstrap_1}. The greenhouse-only ensemble as well as the aerosol-only ensemble exhibit increased variance, signaling that greenhouse and aerosol emissions likely both play a significant role in ENSO's forced response in the full-forcing ensemble. Interestingly, the influence of greenhouse gasses and aerosols are non-conflicting, in contrast with previous studies that show opposite effects of greenhouse gas and aerosol forcing on ENSO \citep{stevenson2017forced}. All the other single forcing ensembles exhibit insignificant differences from the fully-forced ensemble. The biomass burning case shows very small deviations from the fully-forced case, while the ozone ensemble's period of recording is too small to draw meaningful conclusions. However, \citet{nowack2017role} showed that ozone forcing may dampen the effect of greenhouse-forced increases to ENSO amplitude by reducing changes to Pacific sea temperature and the Walker Circulation. The land use/cover case, while it does show large deviations from the fully-forced case, has an ensemble size (5 members) too small to lend any credibility to these changes.

@@

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{../../data/figures/cesm1_sf.pdf}
\caption{\label{fig:bootstrap_1}Difference between fully forced and single forced ensembles derived from the bootstrap process for a) aerosol emissions, b) biomass burning c) greenhouse gasses, d) land use/cover}
\end{figure}

\subsection{Correlation With Ocean Temperature}

Next, we analyzed the correlation between ENSO amplitude and changes to ocean structure in the CESM1 ensembles. To do this we used 4 slices of ocean temperature from the fully-forced, xghg, and xaer ensembles, including a slice averaged along the equator and slices through the western, central, and eastern Pacific basins. We linearly detrend and smooth with a 30-year windowed mean the timeseries at each grid-point and the Niño 3.4 variance. Next, we calculated the Pearson's correlation coefficient between each grid-point and the Niño 3.4 variance timeseries. The correlation coefficients for the equatorial slice are shown in figure \ref{fig:tempdt}.

@@

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{../../data/figures/tempdt.pdf}
\caption{\label{fig:tempdt}Correlation coefficient between detrended and smoothed ocean temperature averaged along the equator and detrended and smoothed Niño 3.4 variance.}
\end{figure}

\subsection{Wavelet Analysis}

Finally, we conduct a wavelet analysis of the Niño 3.4 index for the CESM1 and CESM2 Large Ensembles. Wavelet analysis separates a timeseries into time and frequency domains, displaying the amplitude for each frequency at each time step. The wavelet transform works by scaling a specified ``mother wavelet'' and translating it across the series at each scaling. This method is superior to a windowed Fourier Transform because it has high resolution on both time and frequency domains. First, we subtract the ensemble mean from each member and apply a 1-year windowed mean. Next, we construct a power spectrum of the Niño 3.4 index for each member following the methods of \cite{torrence1998practical} and calculate the ensemble mean of these power spectra, as displayed in figure \ref{fig:wavelet_2}. The majority of ENSO variability is contained in the 2 to 8-year frequency band. This band exhibits steadily increasing wavelet power until the mid 21\textsuperscript{st} century, conforming to previous results of this study. Both ensembles show an increase in the dominant ENSO signal, peaking in the 21\textsuperscript{st} century. The CESM2 ENSO amplitude peaks slightly earlier than the CESM1, in accordance with figure \ref{fig:variance_2}. The CESM2 exhibits a slight decrease in the dominant ENSO period during the 21\textsuperscript{st} century, with the mean power spectrum's ridgeline decreasing from a period of approximately 4 years to 2.5 years. It is unknown whether this period decrease is statistically robust. Finally, the CESM2 has a lower peak wavelet power than the CESM1, in contrast with that shown in figure \ref{fig:variance_2}.

@@

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{../../data/figures/wavelet2.pdf}
\caption{\label{fig:wavelet_1}Ensemble mean wavelet power spectra for both large ensembles.}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{../../data/figures/wavelet3.pdf}
\caption{\label{fig:wavelet_2}Ensemble mean wavelet power records for low and high period ENSO.}
\end{figure}

\section{Conclusion}

@@

\section{Acknowledgments}

>>>>

% Created 2021-10-28 Thu 09:33
% Intended LaTeX compiler: pdflatex
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{natbib}
\author{Benjamin Goldman}
\date{\today}
\title{The Impact of Anthropogenic Forcing on ENSO Amplitude}
\hypersetup{
 pdfauthor={Benjamin Goldman},
 pdftitle={The Impact of Anthropogenic Forcing on ENSO Amplitude},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 27.2 (Org mode 9.5)}, 
 pdflang={English}}
\begin{document}

\maketitle
\begin{abstract}
The El Niño/Southern Oscillation (ENSO) is the dominant mode of interannual climate variability, with substantial associated global socio-economic impacts. Due to their significance, shifts in ENSO under climate change also have the potential to substantially impact human society and natural ecosystems. However, it is currently unclear what effect greenhouse gas (GHG) and industrial aerosol (AER) emissions have on ENSO, as well as what effect these factors have when combined. This study examined transient changes to ENSO variance under a variety of forcing scenarios using the CESM1 and CESM2 Large and Single-Forcing Ensembles. These multi-member ensembles span the historical record (1920-2005) and much of the 21st C (2006-2080 for GHG/AER). A 2000-year pre-industrial (PI) control simulation is used to account for model drift and 20-year running variance of the Niño 3.4 SST index is used as a proxy for ENSO variance. The ensemble mean and standard error of each ensemble was calculated, while the Probability Density Function (PDF) is computed for the PI control simulation to estimate the statistical significance of simulated changes. We calculated the correlation coefficient between ocean temperature in the equatorial Pacific and Niño 3.4 under various forcing conditions, concluding that Pacific stratification likely is tied to changes to ENSO amplitude. We identified significant increases in variance of the Niño 3.4 index under full-forcing conditions during the historical record and attribute these mainly to changes in GHG, with the potential emergence of AER-driven increases in the decades to come.
\end{abstract}

\section{Introduction}
\label{sec:orgfd64463}

@@

\section{Data}
\label{sec:orgf0dfd76}

@@

\section{Methods and Results}
\label{sec:org9a01f67}

\subsection{Niño 3.4 Variance}
\label{sec:org4dc417f}
We estimate 20\textsuperscript{th} and 21\textsuperscript{st} century ENSO amplitude for each model in the fully-forced, single-forced, and PI control CESM1 and CESM2 ensembles using the variance of the Niño 3.4 region of the Pacific ocean (5N-5S, 170W-120W). First, we subtract the ensemble mean from each member to isolate the internal variability for each member. We measure variance on 20-year centered sliding windows. We calculate the multi-model mean for each ensemble, as well as the ensemble standard error. The results of this calculation  for the CESM1 Large and Single Forcing Ensembles are shown in figure \ref{fig:variance_1}.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{../../data/figures/cesm1.pdf}
\caption{\label{fig:variance_1}20-year windowed variance of the Niño 3.4 index, for ensembles a) full-forcing, b) xghg, c) xaer, d) xbmb, e) xlulc. Colored bars represent ensemble standard error.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{../../data/figures/ff_compare.pdf}
\caption{\label{fig:variance_2}20-year windowed variance of the Niño 3.4 index for the CESM1 and CESM2 fully-forced Large Ensembles. Colored bars represent ensemble standard error.}
\end{figure}

The fully-forced ensemble exhibits reduced variance in the mid-late 20\textsuperscript{th} century, below that of the PI control. The most likely explanation for this phenomenon is internal variability. To test the likelihood of this explanation, we test the influence of the Atlantic Multidecadal Oscillation (AMO) and Atlantic Meridional Overturning Current (AMOC) on ENSO in the control simulation. The AMO has been shown to have an influence on ENSO strength and seasonal growth rate \citep{levine2017impact}. We filter the control Niño 3.4 variance data according to the strength of the AMO and AMOC, using records of AMO/C strength in the Climate Variability Diagnostics Package (CVDP) \citep{phillips2014evaluating}. The Probability Distribution Function of Niño 3.4 variance is estimated for AMO/C > 1/2\(\sigma\) and < -1/-2\(\sigma\) using a Kernel Density Estimation. No meaningful differences were found in the distribution of ENSO intensity under any of these conditions (See Supplementary information figure 1).

\subsection{Single Forcing Scenarios and the Bootstrap Process}
\label{sec:org2956178}
We analyze the role of individual factors using the CESM Single Forcing Ensembles. To separate the influence of a single factor from the fully-forced ensemble, we employ a bootstrap test. For each single-forcing ensemble, a single simulation is randomly selected, and the Niño 3.4 20-year variance record is subtracted from that from a randomly selected fully-forced simulation. We repeat this process 1000 times for each ensemble, and then calculate the mean and standard error for each ensemble. These results are shown in figure \ref{fig:bootstrap_1}. The greenhouse-only ensemble as well as the aerosol-only ensemble exhibit increased variance, signaling that greenhouse and aerosol emissions likely both play a significant role in ENSO's forced response in the full-forcing ensemble. Interestingly, the influence of greenhouse gasses and aerosols are non-conflicting, in contrast with previous studies that show opposite effects of greenhouse gas and aerosol forcing on ENSO \citep{stevenson2017forced}. All the other single forcing ensembles exhibit insignificant differences from the fully-forced ensemble. The biomass burning case shows very small deviations from the fully-forced case, while the ozone ensemble's period of recording is too small to draw meaningful conclusions. However, \citet{nowack2017role} showed that ozone forcing may dampen the effect of greenhouse-forced increases to ENSO amplitude by reducing changes to Pacific sea temperature and the Walker Circulation. The land use/cover case, while it does show large deviations from the fully-forced case, has an ensemble size (5 members) too small to lend any credibility to these changes.

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{../../data/figures/cesm1_sf.pdf}
\caption{\label{fig:bootstrap_1}Difference between fully forced and single forced ensembles derived from the bootstrap process for a) aerosol emissions, b) biomass burning c) greenhouse gasses, d) land use/cover}
\end{figure}

@@

\subsection{Correlation With Ocean Temperature}
\label{sec:org1985d36}
Next, we analyzed the correlation between ENSO amplitude and changes to ocean structure in the CESM1 ensembles. To do this we used 4 slices of ocean temperature from the fully-forced, xghg, and xaer ensembles, including a slice averaged along the equator and slices through the western, central, and eastern Pacific basins. We linearly detrend and smooth with a 30-year windowed mean the timeseries at each grid-point and the Niño 3.4 variance. Next, we calculated the Pearson's correlation coefficient between each grid-point and the Niño 3.4 variance timeseries. The correlation coefficients for the equatorial slice are shown in figure \ref{fig:tempdt}.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{../../data/figures/tempdt.pdf}
\caption{\label{fig:tempdt}Correlation coefficient between detrended and smoothed ocean temperature averaged along the equator and detrended and smoothed Niño 3.4 variance.}
\end{figure}

\subsection{Wavelet Analysis}
\label{sec:orgcd04aa3}
Finally, we conduct a wavelet analysis of the Niño 3.4 index for the CESM1 and CESM2 Large Ensembles. Wavelet analysis separates a timeseries into time and frequency domains, displaying the amplitude for each frequency at each time step. The wavelet transform works by scaling a specified ``mother wavelet'' and translating it across the series at each scaling. This method is superior to a windowed Fourier Transform because it has high resolution on both time and frequency domains. First, we subtract the ensemble mean from each member and apply a 1-year windowed mean. Next, we construct a power spectrum of the Niño 3.4 index for each member following the methods of \cite{torrence1998practical} and calculate the ensemble mean of these power spectra, as displayed in figure \ref{fig:wavelet_2}. The majority of ENSO variability is contained in the 2 to 8-year frequency band. This band exhibits steadily increasing wavelet power until the mid 21\textsuperscript{st} century, conforming to previous results of this study. Both ensembles show an increase in the dominant ENSO signal, peaking in the 21\textsuperscript{st} century. The CESM2 ENSO amplitude peaks slightly earlier than the CESM1, in accordance with figure \ref{fig:variance_2}. The CESM2 exhibits a slight decrease in the dominant ENSO period during the 21\textsuperscript{st} century, with the mean power spectrum's ridgeline decreasing from a period of approximately 4 years to 2.5 years. It is unknown whether this period decrease is statistically robust. Finally, the CESM2 has a lower peak wavelet power than the CESM1, in contrast with that shown in figure \ref{fig:variance_2}.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{../../data/figures/wavelet2.pdf}
\caption{\label{fig:wavelet_1}Ensemble mean wavelet power spectra for both large ensembles.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{../../data/figures/wavelet3.pdf}
\caption{\label{fig:wavelet_2}Ensemble mean wavelet power records for low and high period ENSO.}
\end{figure}

\section{Conclusion}
\label{sec:orgf6b6554}

@@

\section{Acknowledgments}
\label{sec:orgb2d1109}
